Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R

### 2017-04-24 meehan_et_al_inla_nmix_jss/article.tex
got Nicole's comments integrated

%% an abstract and keywords
\Abstract{
  Successful management of wildlife populations requires accurate estimates of abundance. Abundance estimates can be confounded by imperfect detection during wildlife surveys. N-mixture models enable quantification of detection probability and often produce abundance estimates that are less biased. The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows Frequentist inference. We show that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \proglang{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.
}
\Keywords{abundance, detection, \proglang{JAGS}, N-mixture model, \proglang{R}, \pkg{R-INLA}, \pkg{unmarked}, wildlife}
%% without formatting
\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}

@@

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}. One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area. If, during repeat visits, the population is assumed to be closed (no immigration, emigration, or reproduction; i.e., static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability. This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.

@@

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for wildlife species that breed annually \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{Binomial}(N_{i,k}, p_{i,k})$. Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates. There are other variations of N-mixture models that accommodate overdispersed counts through use of a Negative Binomial distribution \citep{Kery_Royle_2010}, a Zero-Inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}. Yet other variations account for non-independent detection probabilities through use of a Beta-Binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}. We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for an overview.

@@

There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.  The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a Frequentist framework, with the capacity to accommodate overdispersion and dynamic populations. The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} syntax. Models are analyzed using a Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach. The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists. However, it comes at a cost -- loss of the intuitive inferential framework associated with Bayesian analysis.

@@

As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion was generated and modeled using a Negative Binomial distribution for the count component of the model. The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.b4 = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

@@

\begin{CodeInput}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # number replicate surveys
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.b4 = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{CodeInput}

@@

%% figure 2
\begin{figure}
  \includegraphics[width=\linewidth]{fig2.png}
  \caption{Differences between the estimated posterior mean values and the true input parameter values as a function of the slope coefficient (effect size) for the site-survey-year covariate, x.p.3.  Black circles and lines are output from the model with the site-survey-year covariate, x.p.3, and gray circles and lines are from the model with an averaged site-year covariate, x.p.3.mean.  Parameter name and true input value is given in the strip across the top of each panel.}
  \label{fig:fig2}
\end{figure}
%% end figure 2

@@

The 95\% credible intervals from the \pkg{R-INLA} model overlapped broadly with the 95\% confidence intervals from \pkg{unmarked} model, so parameter estimates were not significantly different from one another (Fig. \ref{fig:fig3}). Also, the parameters with estimates significantly different from zero were the same regardless of technique (Fig. \ref{fig:fig3}). Using both techniques, detection decreased linearly as the season progressed, and abundance decreased as linear functions of increasing forest cover and elevation gain. Parameter estimates and biological conclusions were similar despite the fact that site-survey detection covariates were used for \pkg{unmarked} and site-averaged detection covariates were used for \pkg{R-INLA}. Note that both approaches estimated moderate effects of detection covariates which, according to the results in Example II, would indicate that other parameter estimates were not substantially biased. These results may have been different with a different real dataset, where detection covariates had very strong effects.

@@

%% discussion
\section[Discussion]{Discussion}
The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows Frequentist inference.

@@

The key observation is that both the Poisson and the Binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{Binomial}(y; n, p) = \text{Binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-Binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{Binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{Binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Probability}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{Binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.

>>>>

%% an abstract and keywords
\Abstract{
  Successful management of wildlife populations requires accurate estimates of abundance. Abundance estimates can be confounded by imperfect detection during wildlife surveys. N-mixture models enable quantification of detection probability and often produce abundance estimates that are less biased. The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows frequentist inference. We show that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \proglang{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.
}
\Keywords{abundance, detection, \proglang{JAGS}, N-mixture model, \proglang{R}, \pkg{R-INLA}, \pkg{unmarked}, wildlife}
%% without formatting
\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}

@@

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}. One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area. If, during repeat visits, the population is assumed to be closed (no immigration, emigration, reproduction or mortality; i.e., static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability. This detection probability can be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.

@@

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for wildlife species that breed annually \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \ Y_{i,j,k} | N_{i,k} \sim \text{Binomial}(N_{i,k}, p_{i,k})$. Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates. There are other variations of N-mixture models that accommodate overdispersed counts through use of a Negative Binomial distribution \citep{Kery_Royle_2010}, a Zero-Inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}. Yet other variations account for non-independent detection probabilities through use of a Beta-Binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}. We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for an overview.

@@

There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.  The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a frequentist framework, with the capacity to accommodate overdispersion and dynamic populations. The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} modeling syntax. Models are analyzed using Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach. The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists. However, it comes at a cost -- loss of the intuitive inferential framework associated with Bayesian analysis.

@@

As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion was generated and modeled using a Negative Binomial distribution for the count component of the model. The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.yr = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

@@

\begin{CodeInput}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # number replicate surveys
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.yr = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.yr*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{CodeInput}

@@

%% figure 2
\begin{figure}
  \includegraphics[width=\linewidth]{fig2.png}
  \caption{Differences between the estimated posterior mean values and the true input parameter values as a function of the slope coefficient (effect size) for the site-survey-year covariate, x.p.3. Black circles and lines are from the model with the site-survey-year covariate, x.p.3, and gray circles and lines are from the model with an averaged site-year covariate, x.p.3.mean. Parameter name and true input value is given in the strip across the top of each panel.}
  \label{fig:fig2}
\end{figure}
%% end figure 2

@@

The 95\% credible intervals from the \pkg{R-INLA} model overlapped broadly with the 95\% confidence intervals from the \pkg{unmarked} model, so parameter estimates were not significantly different from one another (Fig. \ref{fig:fig3}). Also, the parameters with estimates significantly different from zero were the same regardless of technique (Fig. \ref{fig:fig3}). Using both techniques, detection decreased linearly as the season progressed, and abundance decreased as linear functions of increasing forest cover and elevation gain. Parameter estimates and biological conclusions were similar despite the fact that site-survey detection covariates were used for \pkg{unmarked} and site-averaged detection covariates were used for \pkg{R-INLA}. Note that both approaches estimated moderate effects of detection covariates which, according to the results in Example II, would indicate that other parameter estimates were not substantially biased. These results may have been different with a different real dataset, where detection covariates had very strong effects.

@@

%% discussion
\section[Discussion]{Discussion}
The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows frequentist inference.

@@

The key observation is that both the Poisson and the Binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{Binomial}(y; n, p) = \text{Binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-Binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{Binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{Binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Probability}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{Binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-likelihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.
